% this file is part of the Data Analysis Recipes project
% Copyright 2012 David W. Hogg

\documentclass[12pt,twoside,pdftex]{article}
\input{./hogg_style}

% header stuff
\renewcommand{\MakeUppercase}[1]{#1}
\pagestyle{myheadings}
\renewcommand{\sectionmark}[1]{\markright{\thesection.~#1}}
\markboth{Extracting information from images}{Introduction}

\begin{document}
\thispagestyle{plain}\raggedbottom
\section*{Data analysis recipes:\\
  Extracting information from images\footnotemark}

\footnotetext{%
  The \notename s begin on page~\pageref{note:first}, including the
  license\note{\label{note:first}%
    Copyright 2012 David W. Hogg (NYU).  Right now this document
    (though publicly available) is NOT FOR DISTRIBUTION.  Eventually,
    and once it is ready, I will release it with the license ``You may
    copy and distribute this document provided that you make no
    changes to it whatsoever''.}
  and the acknowledgements\note{%
    It is a pleasure to thank
      Jo Bovy (IAS),
      Dan Foreman-Mackey (NYU),
      Hans-Walter Rix (MPIA),
      and Dan Weisz (UW)
    for discussions and comments.  This
    research was partially supported by the US National Aeronautics
    and Space Administration and National Science Foundation.}}

\noindent
David~W.~Hogg\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics, New York University}\\
and~\affil{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
%% \\[1ex]
%% A. Nother Author
%% \affil{Oxford University}

\begin{abstract}
Like, so whatever.
\end{abstract}

\section{What is an image?}

... Images are intensity maps, not grids of flux buckets.  The
intensity field is a quantum wave function, which is collapsed at the
detector surface, resulting in photon--electron conversions and
thereby detections.

... What does it mean for an image to be band-limited?

... The ``flux bucket'' image leads to abominations like
\textsl{Drizzle}, where photons in an image are ``poured'' into
``pixel buckets'' in a different coordinate grid.  In Band-limited
imaging, methods for interpolation perform the same task without loss
of resolution but they are only easy to see in the intensity map
formalism.

... Intensity is a conserved quantity; there are many consequences of
this, some of them obvious, some of them non-trivial but I am not sure
which is which.  You can't make the blue sky brighter with normal
transparent optics!  And etc.  Telescopes work by converting position
diversity at fixed direction into direction diversity at fixed
position.  All this happens at fixed (or actually slightly reduced)
intensity.

... Calibration of images, photometric, astrometric, PSF, noise
properties.  self-calibration.  What is meant by a world coordinate
system, a camera model, a photometric calibration, a flat-field, bias,
dark, and so-on.  A typical CCD noise model.

\section{Measurements in images}

... Why all measurements can be seen as optimizations.  Why not
optimize the likelihood?  Or even better produce a full likelihood
function or sample a posterior?

... A measurement of something in an image is inference under a model,
or an approximation thereto

... Likelihood functions are noise models.

... What is blind and non-blind deconvolution and how are they done?
Examples?
 
... Deconvolution is what \emph{everyone does always}.  People claim
to be ``against'' deconvolution, but every positional measurement of
every star ever made has been a deconvolution.

... How to make catalogs and astrometric and photometric measurements.

... What is a catalog and what is its relationship to an image?

... Consider photometry -- using matched radial profiles or
``apertures''.

... Note that standard aperture photometry is the fitting of a
\emph{very inappropriate} model with a \emph{wrong} noise model.  It
also doesn't account for varying PSF or anything else.  That said, it
is very insensitive to small-scale PSF and WCS issues.

... A model of a set of images is no harder, in principle or in
practice, than a model of a single image

... How to measure time-dependent phenomena on sources too faint to
see in any one image?

\section{Coadding and non-parametric scene inference}

... ``Coadding'' is only justified in extremely limited circumstances

... Coadding prohibits certain kinds of investigations.

... Coadding leads to a non-stationary PSF.

... Many images of the same scene give you more information about the scene than any one image

... How do we capitalize on these sets of images?  When they are taken in different bands and under different PSFs?

... Need to face interpolation; how this works in band-limited data.

\section{Advanced topics}

... What about 2-d spectroscopy?

... What about more complicated objects, like light-field cameras and
imaging spectrographs?

... What about interferometry?  Michelson vs Fizeau.

\clearpage
\markright{Notes}\theendnotes

\clearpage
\begin{thebibliography}{}\markright{References}\raggedright
\bibitem[Hogg \etal(2010a)]{straightline}
  Hogg,~D.~W., Bovy,~J., \& Lang,~D., 2010a,
  ``Data analysis recipes:\ Fitting a model to data'', arXiv:1008.4686
\end{thebibliography}

\end{document}
